% =============================================================================
% RCS --- Formal Definitions (IEEE Conference Format)
% Recursive Controlled Systems: Core mathematical framework
% Compile: tectonic rcs-definitions-ieee.tex
% =============================================================================
\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

% --- Encoding and fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% --- Colors (must load before hyperref) ---
\usepackage{xcolor}

% --- Hyperlinks ---
\usepackage[colorlinks=true,linkcolor=blue!70!black,citecolor=blue!70!black,urlcolor=blue!70!black]{hyperref}

% --- Tables ---
\usepackage{booktabs}
\usepackage{array}

% --- Micro-typography ---
\usepackage{microtype}

% --- Algorithms (if needed later) ---
\usepackage{cite}

% --- Shared RCS preamble (math, theorems, macros) ---
% \SHAREDLATEX points at the shared LaTeX assets dir. Each paper in
% papers/pN-.../ resolves it relative to its own location; preamble.tex
% uses the macro to reference its own includes (parameters.tex, etc.).
\def\SHAREDLATEX{../../latex}
\input{\SHAREDLATEX/preamble}

% =============================================================================
\title{Recursive Controlled Systems:\\
A Self-Similar Formalization for\\Autonomous AI Agents}

\author{
  \IEEEauthorblockN{Carlos D. Escobar-Valbuena}
  \IEEEauthorblockA{
    Maestr\'ia en Inteligencia Artificial (MAIA)\\
    Universidad de los Andes\\
    Bogot\'a, Colombia\\
    carlosdavidescobar@gmail.com
  }
}

% =============================================================================
\begin{document}

\maketitle

\begin{abstract}
We introduce the Recursive Controlled System (RCS), a control-theoretic
\emph{type signature} $\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$
whose controller~$\ctrl$ is itself an RCS at the next hierarchical
level.  The formalization is agnostic with respect to the controller
class: classical regulators, latent world models, JEPA-family
predictors, and LLM-based policies each enter as instantiations that
discharge the same assumptions on their own state space.  We define the
homeostatic drive as a common notation covering Lyapunov, reward, and
free-energy interpretations in their respective regimes.  Extending the
stability budget of Eslami \& Yu~\cite{eslami2026control} to
finite-depth hierarchies, we prove by induction on depth~$N$ that if
the per-level stability budget is positive and a time-scale-separation
bound holds, the composite state decays exponentially at rate
$\omega_N = \min_i \stab_i$.  An instantiation catalogue illustrates
how LQR, Dreamer, V-JEPA~2, and LLM controllers populate the type
signature.
\end{abstract}

\begin{IEEEkeywords}
  control systems, Lyapunov stability, multi-agent systems, autonomous agents,
  homeostatic regulation, switched systems, AI safety
\end{IEEEkeywords}

% =============================================================================
\section{Introduction}
% =============================================================================

Autonomous AI agents increasingly operate as controllers of complex
systems, from energy microgrids to software infrastructure.  Yet the
agent itself---its internal state, cognitive resources, and economic
constraints---is a dynamical system that must also be regulated.  This
creates a recursive structure: one controller regulates a plant, a
meta-controller regulates the first controller, and a governance layer
constrains the meta-controller.

\paragraph{Framework, not instantiation.}
We formalize this structure \emph{agnostically with respect to the
controller implementation}.  The Recursive Controlled System (RCS) is a
7-tuple $\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$---a
\emph{type signature} over which stability theorems
operate---not a commitment to any particular plant or policy class.
Concrete controllers (classical LQR
regulators~\cite{khalil2002nonlinear}, Dreamer-style latent world
models~\cite{hafner2025dreamerv3}, JEPA-family
predictors~\cite{assran2025vjepa2}, and LLM-based
policies~\cite{bhargava2023magic,soatto2023taming}) are
\emph{instantiations} of the type signature, each entering by
discharging the per-level assumptions in its own regime.
Section~\ref{sec:instantiations} catalogues four representative
instantiations with their stability witnesses.

\paragraph{Open problem this paper addresses.}
The closest published articulation of the same ambition is LeCun's
position paper~\cite{lecun2022path}, which sketches a six-module
autonomous agent architecture with an intrinsic cost module whose
stability is \emph{assumed} rather than derived, and an informal
hierarchical JEPA (H-JEPA) whose composite stability is not stated as a
theorem.  Our finite-depth induction (Theorem~\ref{thm:recursive})
closes this gap: given per-level assumptions that modern world
models~\cite{hafner2025dreamerv3,assran2025vjepa2}, classical
regulators~\cite{khalil2002nonlinear}, and LLM controller
architectures~\cite{bhargava2023magic,soatto2023taming,wang2026agentspec}
are each known to satisfy on their respective state spaces, we obtain
composite exponential stability for any hierarchy of finite depth.

\paragraph{Contributions.}
\begin{enumerate}
  \item The RCS definition as a recursive type signature
        (Section~\ref{sec:rcs}).
  \item The homeostatic drive as a common notation covering three
        regime-restricted interpretations (Section~\ref{sec:drive}).
  \item A recursive stability budget extending Eslami \& Yu's
        Theorem~2~\cite{eslami2026control} to finite-depth hierarchies,
        proven by induction on depth (Section~\ref{sec:stability}).
  \item An instantiation catalogue showing how LQR, Dreamer, V-JEPA~2,
        and LLM-as-controller each discharge the assumptions
        (Section~\ref{sec:instantiations}).
  \item The EGRI meta-controller as an instance of~$\rcs_2$, with two
        of its Five Laws admitting derived formal content and the
        remaining three recast as design principles
        (Section~\ref{sec:egri}).
\end{enumerate}

% =============================================================================
\section{Preliminaries}
\label{sec:prelim}
% =============================================================================

We recall standard definitions following
Khalil~\cite{khalil2002nonlinear} and
Liberzon~\cite{liberzon2003switching}.

\begin{definition}[Discrete-time dynamical system]
A \emph{discrete-time dynamical system} is a triple $(\X, \U, \dyn)$ where
$\X \subseteq \setR^n$, $\U \subseteq \setR^m$, and
$\dyn : \X \times \U \to \X$:
\begin{equation}
  x_{k+1} = \dyn(x_k, u_k).
\end{equation}
\end{definition}

\begin{definition}[Average dwell time~{\cite{hespanha1999stability}}]
\label{def:dwell-time}
A switching signal $\mode(t)$ has \emph{average dwell time} $\taua$ if
$N_\mode(t, t{+}T) \leq N_0 + T/\taua$ for all $t, T \geq 0$.
\end{definition}

\begin{remark}[Standing-max convention for $N_0$]
\label{rem:N0-standing-max}
In the recursive setting (Theorem~\ref{thm:recursive}), we write
$N_0 := \max_i N_{0,i}$, where $N_{0,i}$ is the chatter bound of
level~$i$'s switching signal; (H5) is read under this convention.
\end{remark}

% =============================================================================
\section{Controlled System}
\label{sec:controlled-system}
% =============================================================================

\begin{definition}[Controlled System]
\label{def:cs}
A \emph{Controlled System} is a 7-tuple
$\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$
where $\X$ is the state space, $\Y$ the observation space, $\U$ the
control input space, $\dyn : \X \times \U \to \X$ the dynamics,
$\obs : \X \to \Y$ the observation map, $\shield : \U \times \X \to \U$
the safety shield, and $\ctrl : \Y^* \to \U$ the controller.

The shield filters unsafe actions via:
\begin{equation}
  u_{\text{safe}} = \shield(\tilde{u}, x)
  = \argmin_{u \in \U_{\text{safe}}(x)} \norm{u - \tilde{u}}^2.
\end{equation}
\end{definition}

% =============================================================================
\section{Recursive Controlled System}
\label{sec:rcs}
% =============================================================================

\begin{definition}[Recursive Controlled System]
\label{def:rcs}
A \emph{Recursive Controlled System} (RCS) is a Controlled System
$\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$ where
$\ctrl = \rcs' = (\X', \Y', \U', \dyn', \obs', \shield', \ctrl')$.

This produces a hierarchy $\Li{0}, \Li{1}, \Li{2}, \ldots$ where level
$\Li{i}$ has controller $\Li{i+1}$, state $\X_{i+1}$ includes the
parameters of $\Li{i}$'s controller, observation $\Y_{i+1}$ derives from
$\Li{i}$'s performance, and control $\U_{i+1}$ modifies $\Li{i}$'s
behavior.
\end{definition}

\begin{remark}[Structural self-similarity]
\label{rem:structural-self-similarity}
The same 7-tuple signature is reused at every level; in practice, the
hierarchy has finite depth $N \leq 4$ (plant, adapter, meta-controller,
governance).  The recursion is notational convenience, and
Theorem~\ref{thm:recursive} proves composite stability by induction on
this finite depth, without invoking a categorical fixed point.
\end{remark}

% =============================================================================
\section{Homeostatic Drive as Lyapunov Function}
\label{sec:drive}
% =============================================================================

\begin{definition}[Homeostatic drive]
\label{def:drive}
At level $\Li{i}$, the \emph{homeostatic drive} is
$\drive{i}(x) = \norm{x - \setpoint{i}}^2$,
where $\setpoint{i}$ is the setpoint at level $i$.
\end{definition}

\begin{proposition}[Three regime-restricted interpretations of $\drive{i}$]
\label{prop:triple}
The drive $\drive{i}$ admits three interpretations, each holding in its
own regime:
(1)~a \textbf{Lyapunov function} certifying asymptotic stability in the
quadratic-$V$ regime where Assumption~\ref{ass:per-level}
holds~\cite{khalil2002nonlinear};
(2)~a \textbf{reward signal} via drive reduction
$r_k = \drive{i}(x_k) - \drive{i}(x_{k+1})$, for bounded physiological
drives with pre-specified setpoints~\cite{keramati2014homeostatic};
(3)~a \textbf{free-energy bound} under linear-Gaussian generative
models with prior preferences at~$\setpoint{i}$, recovering active
inference~\cite{baltieri2019pid,friston2009reinforcement}.
The three regimes overlap only under linear-quadratic dynamics with
pre-specified setpoints; a single notation unifies them, a single
theorem does not.
\end{proposition}

Unlike unbounded reward maximization, $\drive{i}$ saturates at
$\drive{i}(\setpoint{i}) = 0$, which \emph{prevents unbounded resource
acquisition} once equilibrium is
reached~\cite{mineault2024neuroai}; this is narrower than ``safety''
in the specification-gaming or Goodharting sense, which remains
outside the scope of this paper.

% =============================================================================
\section{Recursive Stability Budget}
\label{sec:stability}
% =============================================================================

We extend Eslami \& Yu's stability budget~\cite{eslami2026control} to the
recursive case.

\begin{assumption}[Per-level Lyapunov functions]
\label{ass:per-level}
For each $\Li{i}$, there exists $\lyap{i} : \X_i \to \setR_{\geq 0}$ with
$\underline{\alpha}_i \norm{x}^2 \leq \lyap{i}(x)
\leq \bar{\alpha}_i \norm{x}^2$.
\end{assumption}

\begin{assumption}[Frozen decay]
\label{ass:decay}
With higher-level variables held constant:
$\lyap{i}(x_{k+1}) - \lyap{i}(x_k) \leq -\decay_i \lyap{i}(x_k)$
for some $\decay_i > 0$.
\end{assumption}

\begin{assumption}[Coupling bounds]
\label{ass:coupling}
Adaptation: $|\partial \lyap{i}/\partial\param \cdot \dot\param_i|
\leq L_{\param,i}\rho_i\lyap{i}$.
Design: $|\partial \lyap{i}/\partial d \cdot \dot d_i|
\leq L_{d,i}\eta_i\lyap{i}$.
Delay: $\Delta_\tau \lyap{i} \leq \beta_i\taubar_i\lyap{i}$.
Jumps: $\lyap{i}(x_k^+) \leq \nu_i\lyap{i}(x_k^-)$.
\end{assumption}

\begin{definition}[Stability budget]
\label{def:budget}
The \emph{stability budget} at level $\Li{i}$ is:
\begin{equation}
  \boxed{
  \stab_i = \decay_i - L_{\param,i}\rho_i - L_{d,i}\eta_i
  - \beta_i\taubar_i - \frac{\ln\nu_i}{\taua[i]}
  }
  \label{eq:budget}
\end{equation}
\end{definition}

\paragraph{Standing hypotheses.}
For a finite RCS hierarchy $\rcs_0, \ldots, \rcs_N$ with
$\ctrl_i = \rcs_{i+1}$ for $i < N$, take the following as standing:
\textbf{(H1)}~per-level Lyapunov, Assumption~\ref{ass:per-level};
\textbf{(H2)}~frozen decay $\decay_i > 0$, Assumption~\ref{ass:decay};
\textbf{(H3)}~adaptation coupling with $L_{\param,i}, \rho_i$ (vacuous
at $i = N$);
\textbf{(H4)}~design coupling with $L_{d,i}, \eta_i$ (vacuous at $i = N$);
\textbf{(H5)}~delay, jump, and dwell parameters
$\beta_i, \taubar_i, \nu_i, \taua[i]$,
Assumption~\ref{ass:coupling};
\textbf{(H6)}~time-scale separation: $\exists\, c > 1$ such that
$\taua[i+1] \geq c \cdot \taua[i]$ for every $i < N$;
\textbf{(H7)}~positive budget: $\stab_i > 0$ for every $i$.

\begin{theorem}[Recursive stability, inductive form]
\label{thm:recursive}
Under (H1)--(H7), any finite RCS hierarchy of depth $N \geq 0$ admits a
composite Lyapunov function
$\lyap{\leq N}: \prod_{i=0}^{N} \X_i \to \setR_{\geq 0}$ and scalars
$K_N > 0$, $\omega_N > 0$ with
\begin{align}
  \lyap{\leq N}(\chi(t))
    &\leq K_N\, e^{-\omega_N t}\, \lyap{\leq N}(\chi(0)), \\
  \omega_N &= \min_{0 \leq i \leq N} \stab_i > 0.
\end{align}
\end{theorem}

\begin{proof}
Let $P(N)$ denote the claim for depth $\leq N$.  We induct on~$N$.

\emph{Base case ($N=0$).}  With (H3) and (H4) vacuous at the top, (H1)
and (H2) at $i=0$ combined with (H5) and the Hespanha--Morse
dwell-time formula~\cite{hespanha1999stability} yield
$\lyap{0}(t) \leq \nu_0^{N_0} e^{-\stab_0 t} \lyap{0}(0)$.
Set $\lyap{\leq 0} := \lyap{0}$, $\omega_0 := \stab_0 > 0$ by (H7).

\emph{Inductive step.}  Given $P(n)$, consider depth $n+1$.  Apply
$P(n)$ to the sub-hierarchy $\rcs_0, \ldots, \rcs_n$, with $\rcs_{n+1}$
supplying the quasi-static adaptation/design perturbations to level~$n$
absorbed into $\stab_n$ via (H3)--(H4); this gives $\lyap{\leq n}$ at
rate $\omega_n = \min_{i \leq n} \stab_i$.  At level~$n+1$, (H1), (H2),
and (H5) with vacuous (H3)--(H4) yield
$\lyap{n+1}(t) \leq \nu_{n+1}^{N_0} e^{-\stab_{n+1} t}
\lyap{n+1}(0)$.  By (H6), $\taua[n+1] \geq c\, \taua[n]$ with $c > 1$
places level~$n+1$ strictly slower than the sub-hierarchy, which
discharges the time-scale hypothesis of Tikhonov's singular-perturbation
theorem (Khalil~\cite{khalil2002nonlinear}~Thm.~11.4 /
Liberzon~\cite{liberzon2003switching}~Thm.~3.1).  Choosing
$\alpha_{n+1} > 0$ per the standard Tikhonov construction and setting
\begin{equation}
  \lyap{\leq n+1}(\chi)
    := \lyap{\leq n}(x_0,\ldots,x_n) + \alpha_{n+1}\lyap{n+1}(x_{n+1}),
\end{equation}
the Tikhonov bound yields
$\Delta \lyap{\leq n+1} \leq -\min(\omega_n, \stab_{n+1})
\lyap{\leq n+1}$, giving
$\omega_{n+1} = \min(\omega_n, \stab_{n+1})
= \min_{i \leq n+1} \stab_i$ which is positive by (H7).
\end{proof}

\begin{remark}[Orphan check]
\label{rem:orphan-check}
Every standing hypothesis is invoked at least once: (H1), (H2), (H5)
in the base case and the top-level bound; (H3), (H4) in the coupling
step where they enter $\stab_n$ through the inductive hypothesis;
(H6) in the Tikhonov invocation; (H7) in the positivity of
$\omega_N$.
\end{remark}

% =============================================================================
\section{EGRI as Level~2 Meta-Controller}
\label{sec:egri}
% =============================================================================

The Evaluator-Governed Recursive Improvement (EGRI) framework instantiates
$\rcs_2$ with artifact state as $\X_2$, evaluator scores as $\Y_2$, mutation
operators as $\U_2$, and budget closure + rollback as $\shield_2$.

\paragraph{Two laws admit formal content.}  Two of the Five Laws of
EGRI follow directly from the stability budget:
\begin{law}[Mutation--evaluation proportionality]
\label{law:capacity}
$|\M| \leq \text{capacity}(\score)$ (Ashby~\cite{ashby1956introduction}).
\end{law}
\begin{law}[Budget closure]
\label{law:budget}
$\budget(t_{k+1}) < \budget(t_k)$; $\budget = 0 \Rightarrow$ halt.
\end{law}
\paragraph{Three design principles.}  The remaining three are
\emph{design principles}: policy statements that shape the
meta-controller but do not follow from the stability budget and are
not invoked in any proof.
\textbf{(DP1)~Evaluator supremacy:} loop safety is bounded by evaluator
quality.
\textbf{(DP2)~Evaluator immutability:} never mutate evaluator and
artifact in the same trial; the analogy with the LQG separation
principle is motivating but not derived for arbitrary artifacts.
\textbf{(DP3)~Rollback guarantee:} promoted states are required, by
design, to form a forward-invariant set under the rollback operation.

\begin{proposition}[EGRI--stability coupling]
For $\stab_1 > 0$, the EGRI mutation magnitude must satisfy:
\begin{equation}
  \eta_1 < \frac{\decay_1 - L_{\param,1}\rho_1 - \beta_1\taubar_1
  - \ln\nu_1/\taua[1]}{L_{d,1}}.
\end{equation}
\end{proposition}

% =============================================================================
\section{Observer: Fold as Sufficient Statistic}
\label{sec:observer}
% =============================================================================

\begin{definition}[Fold observer]
\label{def:fold}
At level $\Li{1}$, the \emph{fold observer} computes
$\homostate_{k+1} = \text{fold}(\homostate_k, y_k)$---a Mealy
accumulator over the event stream.  The analogy with a Luenberger
observer (with $A = I$ and full gain on measurements) holds only when
plant dynamics are slower than the folding rate; outside that regime,
the Luenberger framing is notational rather than technical, and no
plant-prediction cancellation is performed.
\end{definition}

\begin{proposition}[Sufficiency]
If $\ctrl_1(y_0, \ldots, y_k) = \ctrl_1(\homostate_k)$, then $\homostate_k$
is a sufficient statistic for regulation decisions.
\end{proposition}

Self-observation consumes resources, adding a reflexive delay
$\taubar_{\text{self}}$ to the budget.  Hysteresis gates (deadband + dwell
time) prevent self-referential oscillation.

% =============================================================================
\section{Multi-Agent Extension}
\label{sec:fleet}
% =============================================================================

A fleet of $N$ agents composes RCS instances over a communication graph
$G$: $\rcs_{\text{fleet}} = \bigotimes_G \rcs^{(j)}$.  Coordination uses
population dynamics~\cite{quijano2017population}, and fleet performance under
disruption is measured via cooperative
resilience~\cite{chacon2025cooperative}:
\begin{equation}
  J_{jl} = \frac{t_i + F_{jl}\Delta t_f + G_{jl}\Delta t_r}
  {t_i + \Delta t_f + \Delta t_r}.
\end{equation}

% =============================================================================
\section{Instantiation Catalogue}
\label{sec:instantiations}
% =============================================================================

The RCS 7-tuple is a \emph{type signature}.  Each concrete controller
populates the signature by discharging the per-level assumptions
(H1)--(H5) on its own state space.  Table~\ref{tab:instantiations}
lists four representative instantiations with the published result
that supplies the per-level~$\decay_i > 0$ required by (H2).  The
table is not exhaustive: additional instantiations (PID-style
activation steering, CBF-QP over continuous plants, EGRI as $\rcs_2$)
can be added by identifying their state space, dynamics, shield, and
per-level decay witness.

\begin{table}[h]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}p{1.0cm}p{1.1cm}p{1.4cm}p{1.3cm}p{2.4cm}@{}}
\toprule
\textbf{Instance} & \textbf{$\X$} & \textbf{$\dyn$} & \textbf{$\shield$} & \textbf{$\decay_i$ witness}\\
\midrule
LQR / classical
  & $\setR^n$
  & $Ax+Bu$
  & CBF~\cite{ames2019cbf}
  & Riccati / LMI; quadratic~$V$~\cite{khalil2002nonlinear}\\
\addlinespace
Dreamer actor--critic
  & Latent $z$
  & RSSM (Lip.)
  & Barrier
  & Spectral bound + symlog/KL~\cite{hafner2025dreamerv3}\\
\addlinespace
V-JEPA~2
  & Latent $\hat z$
  & JEPA + CEM
  & Intrinsic cost
  & Horizon-$T$ pred.\ error~\cite{assran2025vjepa2}\\
\addlinespace
LLM ctrl.
  & Context (shift-grow)
  & Stochastic kernel~\cite{bhargava2023magic}
  & AgentSpec~\cite{wang2026agentspec}
  & $\sigma_{\max}$ + lift~\cite{soatto2023taming}\\
\bottomrule
\end{tabular}
\caption{Four instantiations of the RCS type signature.  The
decay-witness column supplies (H2); per-level (H1, H3--H5) follow from
each row's state-space conventions; the global conditions (H6)--(H7)
are imposed on the composed hierarchy.  When (H1, H3--H7) also hold,
Theorem~\ref{thm:recursive} composes levels into a finite-depth
hierarchy with $\omega_N > 0$.}
\label{tab:instantiations}
\end{table}

Two observations.  First, the LLM row discharges (H1)--(H2) in the
meaning-space quotient of~\cite{soatto2023taming}, not in the raw
token space; the attention spectral bound
of~\cite{bhargava2023magic} provides the contraction rate on the
lifted state.  Second, the four rows use four different mathematical
frameworks (classical Lyapunov, learned latent dynamics, energy-based
world models, discrete stochastic kernels) yet all fit the same
7-tuple signature---which is the point of the type-signature framing.

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

We introduced the Recursive Controlled System, a self-similar formalization
where the same 7-tuple structure applies at every level of the agent
hierarchy.  The recursive stability budget~\eqref{eq:budget} provides
quantitative conditions for composite stability, unifying adaptation rate,
switching frequency, delay, and design evolution costs.  The homeostatic
drive admits a \emph{regime-restricted triple interpretation} as
Lyapunov function, reward signal, and negative free energy
(Proposition~\ref{prop:triple}); each reading holds under stated
conditions, not universally.  Of EGRI's five statements, two (capacity,
budget closure) map to stability inequalities used in the budget, while
the remaining three (DP1--DP3) are design principles that shape the
meta-controller but do not follow from the budget and are not invoked
in any proof.

Future work includes full proofs under weaker time-scale separation
assumptions, empirical validation on the Life Agent OS~\cite{eslami2026control},
and extension to fleet cooperative resilience in Colombian non-interconnected
energy zones.

% =============================================================================
% BIBLIOGRAPHY
% =============================================================================
\bibliographystyle{IEEEtran}
\bibliography{\SHAREDLATEX/references}

\end{document}
