% =============================================================================
% RCS --- Formal Definitions (IEEE Conference Format)
% Recursive Controlled Systems: Core mathematical framework
% Compile: tectonic rcs-definitions-ieee.tex
% =============================================================================
\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

% --- Encoding and fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% --- Colors (must load before hyperref) ---
\usepackage{xcolor}

% --- Hyperlinks ---
\usepackage[colorlinks=true,linkcolor=blue!70!black,citecolor=blue!70!black,urlcolor=blue!70!black]{hyperref}

% --- Tables ---
\usepackage{booktabs}
\usepackage{array}

% --- Micro-typography ---
\usepackage{microtype}

% --- Algorithms (if needed later) ---
\usepackage{cite}

% --- Shared RCS preamble (math, theorems, macros) ---
\input{preamble}

% =============================================================================
\title{Recursive Controlled Systems:\\
A Self-Similar Formalization for\\Autonomous AI Agents}

\author{
  \IEEEauthorblockN{Carlos D. Escobar-Valbuena}
  \IEEEauthorblockA{
    Maestr\'ia en Inteligencia Artificial (MAIA)\\
    Universidad de los Andes\\
    Bogot\'a, Colombia\\
    carlosdavidescobar@gmail.com
  }
}

% =============================================================================
\begin{document}

\maketitle

\begin{abstract}
We introduce the Recursive Controlled System (RCS), a self-similar
formalization of autonomous AI agents grounded in control theory.  An RCS is a
7-tuple $\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$ whose controller
$\ctrl$ is itself an RCS at the next hierarchical level.  We define the
homeostatic drive as a Lyapunov candidate unifying stability analysis,
reinforcement learning, and active inference.  Extending the stability budget
of Eslami \& Yu~\cite{eslami2026control} to the recursive case, we state
conditions under which the composite multi-level system is exponentially
stable.  The formalization subsumes four major frameworks---Eslami's
agency hierarchy, Ashby's ultrastability, Beer's Viable System Model,
and active inference---under a single recursive definition with quantitative
stability conditions.
\end{abstract}

\begin{IEEEkeywords}
  control systems, Lyapunov stability, multi-agent systems, autonomous agents,
  homeostatic regulation, switched systems, AI safety
\end{IEEEkeywords}

% =============================================================================
\section{Introduction}
% =============================================================================

Autonomous AI agents increasingly operate as controllers of complex systems,
from energy microgrids to software infrastructure.  Yet the agent itself---its
internal state, cognitive resources, and economic constraints---is a dynamical
system that must also be regulated.  This creates a recursive structure: the
agent controls a plant, a regulator controls the agent, an optimizer tunes the
regulator, and a governance layer constrains the optimizer.

We formalize this structure as a \emph{Recursive Controlled System} (RCS): a
7-tuple $\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$ whose controller
$\ctrl$ is itself an RCS at the next hierarchical level.  The same
state-space formalism---state, observation, control, dynamics, safety shield,
controller---applies at every level, producing a self-similar hierarchy.

Our main contributions are:
\begin{enumerate}
  \item The RCS definition with self-similar recursion (Section~\ref{sec:rcs}).
  \item The homeostatic drive as a unified Lyapunov function, reward signal,
        and free energy bound (Section~\ref{sec:drive}).
  \item A recursive stability budget extending Eslami \& Yu's
        Theorem~2~\cite{eslami2026control} to multi-level hierarchies
        (Section~\ref{sec:stability}).
  \item The EGRI meta-controller formalized as Level~2 of the hierarchy,
        with Five Laws mapped to stability conditions
        (Section~\ref{sec:egri}).
\end{enumerate}

% =============================================================================
\section{Preliminaries}
\label{sec:prelim}
% =============================================================================

We recall standard definitions following
Khalil~\cite{khalil2002nonlinear} and
Liberzon~\cite{liberzon2003switching}.

\begin{definition}[Discrete-time dynamical system]
A \emph{discrete-time dynamical system} is a triple $(\X, \U, \dyn)$ where
$\X \subseteq \setR^n$, $\U \subseteq \setR^m$, and
$\dyn : \X \times \U \to \X$:
\begin{equation}
  x_{k+1} = \dyn(x_k, u_k).
\end{equation}
\end{definition}

\begin{definition}[Average dwell time~{\cite{hespanha1999stability}}]
\label{def:dwell-time}
A switching signal $\mode(t)$ has \emph{average dwell time} $\taua$ if
$N_\mode(t, t{+}T) \leq N_0 + T/\taua$ for all $t, T \geq 0$.
\end{definition}

% =============================================================================
\section{Controlled System}
\label{sec:controlled-system}
% =============================================================================

\begin{definition}[Controlled System]
\label{def:cs}
A \emph{Controlled System} is a 7-tuple
$\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$
where $\X$ is the state space, $\Y$ the observation space, $\U$ the
control input space, $\dyn : \X \times \U \to \X$ the dynamics,
$\obs : \X \to \Y$ the observation map, $\shield : \U \times \X \to \U$
the safety shield, and $\ctrl : \Y^* \to \U$ the controller.

The shield filters unsafe actions via:
\begin{equation}
  u_{\text{safe}} = \shield(\tilde{u}, x)
  = \argmin_{u \in \U_{\text{safe}}(x)} \norm{u - \tilde{u}}^2.
\end{equation}
\end{definition}

% =============================================================================
\section{Recursive Controlled System}
\label{sec:rcs}
% =============================================================================

\begin{definition}[Recursive Controlled System]
\label{def:rcs}
A \emph{Recursive Controlled System} (RCS) is a Controlled System
$\rcs = (\X, \Y, \U, \dyn, \obs, \shield, \ctrl)$ where
$\ctrl = \rcs' = (\X', \Y', \U', \dyn', \obs', \shield', \ctrl')$.

This produces a hierarchy $\Li{0}, \Li{1}, \Li{2}, \ldots$ where level
$\Li{i}$ has controller $\Li{i+1}$, state $\X_{i+1}$ includes the
parameters of $\Li{i}$'s controller, observation $\Y_{i+1}$ derives from
$\Li{i}$'s performance, and control $\U_{i+1}$ modifies $\Li{i}$'s behavior.
The hierarchy is a fixed point: $\text{RCS} \cong F(\text{RCS})$.
\end{definition}

% =============================================================================
\section{Homeostatic Drive as Lyapunov Function}
\label{sec:drive}
% =============================================================================

\begin{definition}[Homeostatic drive]
\label{def:drive}
At level $\Li{i}$, the \emph{homeostatic drive} is
$\drive{i}(x) = \norm{x - \setpoint{i}}^2$,
where $\setpoint{i}$ is the setpoint at level $i$.
\end{definition}

\begin{proposition}[Triple equivalence]
\label{prop:triple}
Under appropriate mappings, $\drive{i}$ serves simultaneously as:
(1)~a \textbf{Lyapunov function} certifying asymptotic stability;
(2)~a \textbf{reward signal} via drive reduction $r_k = \drive{i}(x_k) -
\drive{i}(x_{k+1})$~\cite{keramati2014homeostatic};
(3)~a \textbf{free energy bound} under linear generative
models~\cite{baltieri2019pid, friston2009reinforcement}.
\end{proposition}

Unlike unbounded reward maximization, $\drive{i}$ saturates at
$\drive{i}(\setpoint{i}) = 0$, providing inherent safety
guarantees~\cite{mineault2024neuroai}.

% =============================================================================
\section{Recursive Stability Budget}
\label{sec:stability}
% =============================================================================

We extend Eslami \& Yu's stability budget~\cite{eslami2026control} to the
recursive case.

\begin{assumption}[Per-level Lyapunov functions]
\label{ass:per-level}
For each $\Li{i}$, there exists $\lyap{i} : \X_i \to \setR_{\geq 0}$ with
$\underline{\alpha}_i \norm{x}^2 \leq \lyap{i}(x)
\leq \bar{\alpha}_i \norm{x}^2$.
\end{assumption}

\begin{assumption}[Frozen decay]
\label{ass:decay}
With higher-level variables held constant:
$\lyap{i}(x_{k+1}) - \lyap{i}(x_k) \leq -\decay_i \lyap{i}(x_k)$
for some $\decay_i > 0$.
\end{assumption}

\begin{assumption}[Coupling bounds]
\label{ass:coupling}
Adaptation: $|\partial \lyap{i}/\partial\param \cdot \dot\param_i|
\leq L_{\param,i}\rho_i\lyap{i}$.
Design: $|\partial \lyap{i}/\partial d \cdot \dot d_i|
\leq L_{d,i}\eta_i\lyap{i}$.
Delay: $\Delta_\tau \lyap{i} \leq \beta_i\taubar_i\lyap{i}$.
Jumps: $\lyap{i}(x_k^+) \leq \nu_i\lyap{i}(x_k^-)$.
\end{assumption}

\begin{definition}[Stability budget]
\label{def:budget}
The \emph{stability budget} at level $\Li{i}$ is:
\begin{equation}
  \boxed{
  \stab_i = \decay_i - L_{\param,i}\rho_i - L_{d,i}\eta_i
  - \beta_i\taubar_i - \frac{\ln\nu_i}{\taua[i]}
  }
  \label{eq:budget}
\end{equation}
\end{definition}

\begin{theorem}[Recursive stability]
\label{thm:recursive}
Consider an RCS hierarchy $\rcs_0, \ldots, \rcs_N$ satisfying
Assumptions~\ref{ass:per-level}--\ref{ass:coupling} with time-scale
separation between levels.  If $\stab_i > 0$ for all $i$,
then the composite state $\chi = (x_0, \ldots, x_N)$ converges
exponentially:
\begin{equation}
  \norm{\chi(t)} \leq K e^{-\omega t} \norm{\chi(0)},
  \quad \omega = \min_i \stab_i.
\end{equation}
\end{theorem}

\begin{proof}[Proof sketch]
By singular perturbation, each level sees higher levels as quasi-static.
The per-level Lyapunov derivative is
$\Delta\lyap{i} \leq -(\stab_i + \ln\nu_i/\taua[i])\lyap{i}$.
Combining with jump growth and applying the average dwell-time
condition~\cite{hespanha1999stability} yields
$\lyap{i}(T) \leq \nu_i^{N_0} e^{-\stab_i T}\lyap{i}(0)$.
The composite $V = \sum_i \alpha_i\lyap{i}$ decays at rate
$\omega = \min_i \stab_i$.
\end{proof}

% =============================================================================
\section{EGRI as Level~2 Meta-Controller}
\label{sec:egri}
% =============================================================================

The Evaluator-Governed Recursive Improvement (EGRI) framework instantiates
$\rcs_2$ with artifact state as $\X_2$, evaluator scores as $\Y_2$, mutation
operators as $\U_2$, and budget closure + rollback as $\shield_2$.

Five laws govern safe meta-control:
\begin{law}[Evaluator supremacy]
Loop safety $\leq$ evaluator quality.
\end{law}
\begin{law}[Mutation-evaluation proportionality]
$|\M| \leq \text{capacity}(\score)$ (Ashby~\cite{ashby1956introduction}).
\end{law}
\begin{law}[Evaluator immutability]
Never mutate evaluator and artifact in the same trial.
\end{law}
\begin{law}[Budget closure]
$\budget(t_{k+1}) < \budget(t_k)$; $\budget = 0 \Rightarrow$ halt.
\end{law}
\begin{law}[Rollback guarantee]
Promoted states form a forward-invariant set.
\end{law}

\begin{proposition}[EGRI--stability coupling]
For $\stab_1 > 0$, the EGRI mutation magnitude must satisfy:
\begin{equation}
  \eta_1 < \frac{\decay_1 - L_{\param,1}\rho_1 - \beta_1\taubar_1
  - \ln\nu_1/\taua[1]}{L_{d,1}}.
\end{equation}
\end{proposition}

% =============================================================================
\section{Observer: Fold as Sufficient Statistic}
\label{sec:observer}
% =============================================================================

\begin{definition}[Fold observer]
At level $\Li{1}$, the \emph{fold observer} computes
$\homostate_{k+1} = \text{fold}(\homostate_k, y_k)$---a Mealy machine
equivalent to a Luenberger observer with $A = I$.
\end{definition}

\begin{proposition}[Sufficiency]
If $\ctrl_1(y_0, \ldots, y_k) = \ctrl_1(\homostate_k)$, then $\homostate_k$
is a sufficient statistic for regulation decisions.
\end{proposition}

Self-observation consumes resources, adding a reflexive delay
$\taubar_{\text{self}}$ to the budget.  Hysteresis gates (deadband + dwell
time) prevent self-referential oscillation.

% =============================================================================
\section{Multi-Agent Extension}
\label{sec:fleet}
% =============================================================================

A fleet of $N$ agents composes RCS instances over a communication graph
$G$: $\rcs_{\text{fleet}} = \bigotimes_G \rcs^{(j)}$.  Coordination uses
population dynamics~\cite{quijano2017population}, and fleet performance under
disruption is measured via cooperative
resilience~\cite{chacon2025cooperative}:
\begin{equation}
  J_{jl} = \frac{t_i + F_{jl}\Delta t_f + G_{jl}\Delta t_r}
  {t_i + \Delta t_f + \Delta t_r}.
\end{equation}

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

We introduced the Recursive Controlled System, a self-similar formalization
where the same 7-tuple structure applies at every level of the agent
hierarchy.  The recursive stability budget~\eqref{eq:budget} provides
quantitative conditions for composite stability, unifying adaptation rate,
switching frequency, delay, and design evolution costs.  The homeostatic
drive serves as a triple-purpose Lyapunov function, and the EGRI meta-controller's
Five Laws map to necessary stability conditions.

Future work includes full proofs under weaker time-scale separation
assumptions, empirical validation on the Life Agent OS~\cite{eslami2026control},
and extension to fleet cooperative resilience in Colombian non-interconnected
energy zones.

% =============================================================================
% BIBLIOGRAPHY
% =============================================================================
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
